Fixing Vulnerabilities Hinders Maintainability

### 2019-02-02 paper/ICPC19/icpc19.tex

Software quality is important because it is ultimately related to the overall
cost for developing, extending and maintaing software applications. Software
quality characteristics include, but are not limited to, functional correctness,
reliability, usability, maintainability, and security. Security is an important
non-functional requirement during the development of software systems.  However,
there is still a considerable amount of zero-day vulnerabilities being
discovered, and fixed, almost weekly as disclosed by the Zero Day Initiative
website\footnote{Zero Day Initiative website available  at
https://www.zerodayinitiative.com/advisories/published/ (Accessed on \today{})}.
As an example, in the last week of January 2019, a severe vulnerability in Apple
FaceTime was
revealed\footnote{https://thehackernews.com/2019/01/apple-facetime-privacy-hack.html
(Accessed on \today{})}. Callers could see and hear others without them picking up the call. These vulnerabilities are gateways for attackers to
penetrate systems and exploit their resources.

\Sof{Models}

In 2011, the International Organization for Standardization (ISO) issued an
update for software product quality ISO/IEC 25010~\cite{iso:2011} considering
\emph{Security} as one of the main software product quality characteristics.
Unmaintainable code is hard to test, analyse and re-use, and thus more difficult
to find vulnerabilities and defects. Therefore, writing maitainable code
prevents the introduction of vulnerabilities in the code. In this work, we are
concerned about the impact that fixing vulnerabilities may have in the
maintainability of open-source software. Are developers affecting software
maintainability when trying to improve software safety? We suspect that some of
those fixes may have a negative impact in the software maintainability and
possibly even be the cause of the introduction of new vulnerabilities.

ISO describes software maintainability as ``the degree of effectiveness and
efficiency with which a software product or system can be modified to improve
it, correct it or adapt it to changes in environment, and in requirements'' on
software quality ISO/IEC 25010. As ISO does not provide any specific guidelines or
formula to calculate maintainability, we resort on Software Improvement
Group (SIG)’s web-based source code analysis service Better Code Hub
(BCH)~\cite{Visser:2016:OREILLY} to compute the maintainability of a given
project. BCH compiles the ISO/IEC 25010 into a set of 10 guidelines that are the
base of BCH's evaluation. Two examples of guidelines used by BCH are the
\emph{Write Simple Units of Code} based on the McCabe Complexity~\cite{1702388}
and \emph{Keep your Codebase Small} based on the idea that smaller codebases are
easier to maintain and less sensitive to defects~\cite{Visser:2016:OREILLY}.
Compliance to the guidelines is derived from the SIG's
industry benchmark which, according to official sources, consists of over 8
billion lines of code in more than 180 different technologies. SIG analyzes
around 15 million lines of code, per week.

The particular subject of this paper is to explore whether there is a trade-off
between refactoring source code to fix identified vulnerabilities while keeping
(or even improving) the maintainability of the software system. In other words,
we are interested in answering the following overarching question

\begin{framed}
\noindent
Does improving security have a negative impact on code
maintainability?
\end{framed}

\noindent
In this paper, we present the results of our analysis on the maintainability of
607 security changes collected from open-source software, from different
programming languages.

The main contributions of the present work is:
\begin{itemize}
	\item An empirical study on the impact of security changes on software
	maintainability and what patterns need more attention.

  % \item An analysis of how likely security-related changes need to be
  % maintained, in comparison to regular commits.

@@

This empirical study exhibits proof that changes applied in the codebase to fix
vulnerabilities affect code maintainability. Especially, for refactorings that
address \emph{Broken Authentication and Session Management} issues, \emph{Memory
Leaks} and \emph{Denial-of-Service} attacks. With this study, we intend to
highlight the need for tools and documentation to help developers adopt security
best practices.

@@

As said before, the web-based source code analysis service \emph{Better Code
Hub} (BCH) is used to collect the maintanability reports of the refactorings of
each project. Table \ref{tab:guidelines} presents the 10 guidelines
proposed by BCH's authors for delivering software that is easy to maintain
\cite{Visser:2016:OREILLY}.

All guidelines are computed in the same evaluation by BCH. During each guideline
evaluation, the tool determines the compliance towards one guideline by
establishing limits for the percentage of code allowed to be in each of the $4$
risk severity levels (\emph{low risk}, \emph{medium risk}, \emph{high risk} and
\emph{very high risk}). If the project does not violate the thresholds, then it
is compliant with the guideline. These thresholds are calibrated by BCH using
their own data/experience --- open-source and closed software systems. If a
project is compliant with a guideline, it means that it is at least $65\%$
better than the software used by BCH to calculate the thresholds\footnote{Check
the answer to \emph{How can I adjust the threshold for passing/not passing a
guideline?} at https://bettercodehub.com/docs/faq (Accessed on \today{})}.

@@

\begin{figure}[h]
 	\centering 	\includegraphics[width=0.45\textwidth]{figures/bch_report.png}
 	\caption{Maintainability report of OpenSSL CVE-2014-3506 vulnerability
refactoring for the guideline \emph{Write Simple Units of Code} provided by
\emph{Better Code Hub}. This version of OpenSSL does not comply with the
guideline in the example since the bars do not reach the threshold points. This
example only complies with $\frac{1}{10}$ guidelines (\emph{Write Clean Code}).}
	\label{fig:bchrep}
\end{figure}

\emph{Units} are the smallest groups of code that can be maintained and executed
independently \cite{Visser:2016:OREILLY}. McCabe \cite{1702388} is a software
metric that calculates the number of linearly independent paths of the source
code. BCH uses this metric to measure the number of branch points in the
\emph{Write Simple Units of Code} guideline. The bar represents the top 30 of
units that violates the guideline, sorted by severity, which is indicated by the
colors of the checkboxes. The green bar represents the number of compliant
branch points per unit (\emph{at most 5}), i.e., the units are compliant with
ISO 25010 \cite{iso:2011}. Yellow, orange and red bars represent units that do
not comply with medium (\emph{above 5}), high (\emph{above 10}) and very high
(\emph{above 25}) severity levels. In the bar, there are marks that pinpoint the
compliance thresholds for each severity level. \Sof{@todo: Adress thresholds confusion}

@@

40.2\% of the security refactoring have a negative impact on the maintainability
of open-source software, Figure \ref{fig:secvsreg}. This raises a new tradeoff when
developers need to fix vulnerabilities on their projects because they might be affecting
the software maintainability or even introducing new vulnerabilities. This study exhibits
proof that developers may have to reduce maintainability for the sake of safety. We understand
that developers should be able to produce secure code without affecting the maintainability of
their projects. But there are still some concerns that need to be addressed:
\begin{itemize}
	\item \textbf{Security Frameworks} need to be secure. Unfortunately, there are several
	vulnerabilities that stay undisclosed. Thus, developers should not trust them blindly and
	check defaults.
	\item \textbf{Documentation} of software should provide developers with the best practices to implement safe and maintainable software. Unmaintainable code is difficult to test, analyze and re-use.
	\item \textbf{Proper Authorization Mechanisms} should be designed and implemented. This is
	one of the most complex security features to implement. Some mechanisms might assist authorization maintenance in long run, such as the combination of \texttt{HTTP} filters with \texttt{Spring Security} annotations.
	\item \textbf{Deprecated code} should be avoided by developers. They should integrate tools such as BCH on their projects CI/CD pipelines to help them be more aware of the maintainability of their code and issues (e.g., dead and duplicated code).
	\item\textbf{Programming Languages} should provide new mechanisms to easily implement security patterns without endangering maintainability.

@@

In contrast to the BCH benchmark, this study uses a dataset that only covers open-source software to perform the maintanibility evaluation. Thus, answering these questions to closed commercial applications may be a future work opportunity.

@@

\subsection{Security Maintainability Models}

~\cite{6616351,oswap:2009,common:2009}

@@

This work presents an empirical study on the impact of 607 security refactorings on
the maintainability of 94 open-source projects. We leveraged Better Code Hub reports
of each project version to calculate maintainability based on a model proposed in
previous work.

Security refactorings significantly affect the software maintainability of open-source
software, i.e., developers hinder maintainability when fixing vulnerabilities. We found
this evidence in 40.2\% of the studied cases. Regarding regular commits, it seems that
overall maintainability increases (50.16\%). However, there is still a very considerable amout of
cases that harm maintaniability (41.35\%). This study also exhibits evidence that some patterns
need more attention than others. Broken Autentication issues, Denial-of-Service attacks
and Memory Leaks endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively.

Following the implications of our work, we draw attention for the need of secure security frameworks,
code best practices documentation to produce safe and maintainable code, proper authorization
mechanisms design and implementation, better tools to introduce in the CI/CD pipeline to help software
maintenance and new or improved programming languages to allow developers to easily implement safety in
their software without endangering maintainability.

As future work, this study may be extended in many directions such as analyze which guidelines are
more likely to affect security refactorings; evaluate the impact of programming languages of the
maintainability of each refactoring pattern; investigate if there is any correlation between
software maintainability and projects popularity; expand our methodology with other software quality
properties; and, validate these findings with closed or private software.

>>>>

Software quality is important because it is ultimately related to the overall
cost for developing, extending and maintaing software applications. Software
quality characteristics include, but are not limited to, functional correctness,
reliability, usability, maintainability, and security. Security is an important
non-functional requirement during the development of software systems.  However,
there is still a considerable amount of zero-day vulnerabilities being
discovered, and fixed, almost weekly as disclosed by the Zero Day Initiative
website\footnote{Zero Day Initiative website available  at
https://www.zerodayinitiative.com/advisories/published/ (Accessed on \today{})}. 
These vulnerabilities are gateways for attackers to penetrate 
systems and exploit their resources. 
As an example, in the last week of January 2019, a severe vulnerability in Apple
FaceTime was revealed\footnote{https://thehackernews.com/2019/01/apple-facetime-privacy-hack.html
(Accessed on \today{})} --- the outcome of a design/logical flaw that allowed callers to see and
hear others without them picking up the call.

In 2011, the International Organization for Standardization (ISO) issued an
update for software product quality ISO/IEC 25010~\cite{iso:2011} considering
\emph{Security} as one of the main software product quality characteristics.
Unmaintainable code is hard to test, analyse and re-use, and thus more difficult
to find vulnerabilities and defects. Therefore, writing maintainable code
prevents the introduction of vulnerabilities in the codebases. In this work, we are
concerned about the impact that patching vulnerabilities may have in the
open-source software maintainability. Are developers affecting software
maintainability when trying to improve software safety? We suspect that some of
the patches may have a negative impact in the software maintainability and
possibly even be the cause of the introduction of new vulnerabilities.

ISO describes software maintainability as ``the degree of effectiveness and
efficiency with which a software product or system can be modified to improve
it, correct it or adapt it to changes in environment, and in requirements'' on
software quality ISO/IEC 25010. As ISO does not provide any specific guidelines or
formula to calculate maintainability, we resort on Software Improvement
Group (SIG)'s web-based source code analysis service Better Code Hub
(BCH)~\cite{Visser:2016:OREILLY} to compute the maintainability of a given
project. There are other well-known standards and models that have been proposed to increase software safety: 
Common Criteria~\cite{common:2009} which suffered a lot of criticism regarding the
costs associated and poor technical evaluation; the OWASP 
Application Security Verification Criteria~\cite{oswap:2009} which is focused only
on web applications, and a model proposed by Xu et al. (2013)~\cite{6616351} for 
raiting software security which we believe it was one of the first steps taken
by SIG to introduce security on their maintainability model.
BCH compiles the ISO/IEC 25010 into a set of 10 guidelines that are the
base of BCH's evaluation. Two examples are the
\emph{Write Simple Units of Code} based on the McCabe Complexity~\cite{1702388}
and \emph{Keep your Codebase Small} based on the idea that smaller codebases are
easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}. Compliance to the guidelines is derived from the SIG's
industry benchmark which, according to official sources, consists of over 8
billion lines of code in more than 180 different technologies. SIG analyzes
around 15 million lines of code, per week.

The particular subject of this paper is to explore whether there is a trade-off
between refactoring source code to patch disclosed vulnerabilities while keeping
(or even improving) the maintainability of software systems. Hedegus et al. (2018)~\cite{HEGEDUS2018313}, 
analyzed the difference of maintainability between
groups of refactored elements and non-refactored and conclude that the source
code elements subjected to refactorings had significantly lower maintainability
than the elements not affected.

In this paper, we present the results of our analysis on the maintainability of
607 security changes collected from open-source software, from different
programming languages. The main contributions of the present work is:
\begin{itemize}
	\item An empirical study on the impact of security changes on software
	maintainability and what patterns need more attention.

@@

This empirical study exhibits proof that changes applied in the codebase to fix
vulnerabilities affect code maintainability. Especially, for refactorings that
address \emph{Broken Authentication and Session Management} issues, \emph{Memory
Leaks} and \emph{Denial-of-Service} attacks. There are already several tools that suggest security improvements using static analysis, dynamic analysis, and even artificial intelligence. With this study, we intend to highlight the need for tools to predict the impact of patches on maintainability and documentation
to help developers adopt security best practices~\cite{6311252, 7927935, MESQUIDA201519}.

@@

As said before, the web-based source code analysis service \emph{Better Code
Hub} (BCH) is used to collect the maintanability reports of the refactorings of
each project. Table \ref{tab:guidelines} presents the 10 guidelines
proposed by BCH's authors for delivering software that is easy to maintain
\cite{Visser:2016:OREILLY}. All guidelines are computed in the same evaluation by 
BCH. During each guideline evaluation, the tool determines the compliance towards 
one guideline by establishing limits for the percentage of code allowed to be in 
each of the $4$ risk severity levels (\emph{low risk}, \emph{medium risk}, \emph{high risk} 
and \emph{very high risk}). If the project does not violate the thresholds, then it
is compliant with the guideline. These thresholds are calibrated by BCH using
their own data/experience --- open-source and closed software systems. If a
project is compliant with a guideline, it means that it is at least $65\%$
better than the software used by BCH to calculate the thresholds\footnote{Check
the answer to \emph{How can I adjust the threshold for passing/not passing a
guideline?} at https://bettercodehub.com/docs/faq (Accessed on \today{})}.

@@

\begin{figure}[h]
 	\centering 	\includegraphics[width=0.45\textwidth]{figures/bch_report.png}
 	\caption{Maintainability report of OpenSSL CVE-2014-3506 vulnerability
refactoring for the guideline \emph{Write Simple Units of Code} provided by
\emph{Better Code Hub}. This version of OpenSSL does not comply with the
guideline in the example since the bars do not reach the threshold points. This
example only complies with $\sfrac{1}{10}$ guidelines (\emph{Write Clean Code}).}
	\label{fig:bchrep}
\end{figure}

\emph{Units} are the smallest groups of code that can be maintained and executed
independently \cite{Visser:2016:OREILLY}. McCabe \cite{1702388} is a software
metric that calculates the number of linearly independent paths of the source
code. BCH uses this metric to measure the number of branch points in the
\emph{Write Simple Units of Code} guideline. The bar represents the top 30 of
units that violates the guideline, sorted by severity, which is indicated by the
colors of the checkboxes. The green bar represents the number of compliant
branch points per unit (\emph{at most 5}), i.e., the units are compliant with
ISO 25010 \cite{iso:2011}. Yellow, orange and red bars represent units that do
not comply with medium (\emph{above 5}), high (\emph{above 10}) and very high
(\emph{above 25}) severity levels. In the bar, there are marks that pinpoint the
compliance thresholds for each severity level. If the green mark is somewhere
in the green bar it means it is compliant with a low level of severity.

@@

40.2\% of the security refactoring have a negative impact on the maintainability
of open-source software, Figure \ref{fig:secvsreg}. This raises a new tradeoff when
developers need to fix vulnerabilities on their projects because they might be affecting
the software maintainability or even introducing new vulnerabilities. This study exhibits
proof that developers may have to reduce maintainability for the sake of safety. We understand
that developers should be able to produce secure code without affecting the maintainability of
their projects. But there are still some concerns that need to be addressed:
\begin{itemize}
	\item \textbf{Security Frameworks} need to be secure and user-friendly~\cite{5287006, 7676144}. Unfortunately, 
	there are several vulnerabilities that stay undisclosed. Thus, developers should not 
	trust them blindly and check defaults.
	\item \textbf{Documentation} of software should provide developers with the best practices to implement safe and maintainable software. Unmaintainable code is difficult to test, analyze and re-use.
	\item \textbf{Proper Authorization Mechanisms} should be designed and implemented. This is
	one of the most complex security features to implement. Some mechanisms might assist authorization maintenance in long run, such as the combination of \texttt{HTTP} filters with \texttt{Spring Security} annotations.
	\item \textbf{Deprecated code} should be avoided by developers. They should integrate tools such as BCH on their projects CI/CD pipelines to help them be more aware of the maintainability of their code and issues (e.g., dead and duplicated code).
	\item\textbf{Programming Languages} should provide new mechanisms to easily implement security patterns without endangering maintainability.

@@

In contrast to the BCH benchmark, this study uses a dataset that only covers open-source software to perform the maintanibility evaluation. However, BCH thresholds are calculated from a set of both open and closed sotfware. Different programming languages may require different coding practices to address software safety. The dataset
comprises more commits in C and PHP, i.e., the dataset may not be properly representative of the population regarding to programming languages.

@@

Many studies have investigated the relationship between refactorings and software quality.

Previous work focused on coupling metrics has evaluated the impact of refactorings and
exhibited proof that quantifying effectively the impact of refactoring on maintainability
may help to choose the appropriate refactoring type~\cite{1167822}. In contrast to
this work, Hegedus et al.~\cite{HEGEDUS2018313} did not select particular metrics to
assess the effect of refactorings. Instead, statistical tests were used to find
the metrics that have the potential to change significantly after refactorings. 
The differences of maintainability between groups of refactored elements and 
non-refactored were analyzed. And, the authors conclude that the source
code elements subjected to refactorings had significantly lower maintainability
than the elements not affected. Studying the evolution of maintaninability issues during the development of
Android apps, Malavolta et al. (2018)~\cite{8530041} discovered that maintainability 
decreases over time. The present work is based on the 10 guidelines proposed
by the Software Improvement Group (SIG) and focuses solely on evaluating
the impact of patching vulnerabilities on software maintainability.

Researchers investigated the relationship between design patterns and
software maintainability~\cite{10.1007/978-3-642-35267-6-18}. 300 revisions
of \texttt{JHotDraw} were analyzed to conclude that every introduced pattern
instance improved software maintainability. However, other studies show that the 
use of design patterns may introduce maintainability issues into software~\cite{4493325}.
The present work studies how security patterns influence maintainability
for open-source software.

%
% Previous work studied the effect of programming languages in the quality of
% code found~\cite{ray2014large}. It was found that language design has a
% significant yet moderate impact on software quality. The authors have used the
% number of defects as a construct of software quality. Our work analyzes
% software quality in terms of code maintainability to study how it is affected
% by energy efficiency-code changes.

@@

This work presents an empirical study on the impact of 607 security refactorings on
the maintainability of 94 open-source projects. We leveraged Better Code Hub reports
of each project version to calculate maintainability based on a model proposed in 
previous work. Security refactorings significantly affect the software maintainability of open-source 
software, i.e., developers hinder maintainability when fixing vulnerabilities. We found
this evidence in 40.2\% of the studied cases. Regarding regular commits, it seems that
overall maintainability increases (50.16\%). However, there is still a very considerable amout of
cases that harm maintaniability (41.35\%).

This study also exhibits evidence that some patterns
need more attention than others. Broken Autentication issues, Denial-of-Service attacks 
and Memory Leaks endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively. Following the implications of our work, we draw attention for the need of secure security frameworks,
code best practices documentation to produce safe and maintainable code, proper authorization 
mechanisms design and implementation, better tools to introduce in the CI/CD pipeline to help software
maintenance and new or improved programming languages to allow developers to easily implement safety in
their software without endangering maintainability.

As future work, this study may be extended in many directions such as analyze which guidelines are
more likely to affect security refactorings; evaluate the impact of programming languages of the
maintainability of each refactoring pattern; investigate if there is any correlation between
software maintainability and projects popularity; expand our methodology with other software quality
properties; validate these findings with closed or private software; and, expand this analysis
to other quality standards.
